Spatial Investigation and Exploration of Factors influencing HIV-related Mortality

### 2023-04-08 code/analysis_code/Race/p1_EDA_race.Rmd
fixed rendering

race_df %>% 
  ggplot() +
  geom_histogram(aes(x = `Ratio Observed/ Expected`, fill = Race), bins = 10)
shapiro.test(race_df$`Ratio Observed/ Expected`)
```
Both histograms show a positively skewed distribution.

@@

>>>>

race_df %>% 
  ggplot() +
  geom_histogram(aes(x = `Ratio Observed/ Expected`, fill = Race), bins = 10)
shapiro.test(race_df$`Ratio Observed/ Expected`)
ggsave(here("results/raw_roe_hist.png"), units="in", width=10, height=7, dpi=300)
```
Both histograms show a positively skewed distribution.

@@

ggsave(here("results/box_hist_facet.png"), units="in", width=10, height=7, dpi=300)

race_df_ROE %>%
  mutate(BOX = box_cox(`Ratio Observed/ Expected`, lambda_ROE)) %>%
  ggplot() +
  geom_histogram(aes(x= BOX, fill = Race), bins = 10) +
  theme_classic() +
  theme(legend.position = "none") +
  labs(x = "Box Cox Transformation of Observed/ Expected HIV Mortality Ratio", y = "Count")

### 2023-04-08 code/analysis_code/Race/p4_pulling_model_fit.Rmd
fixed rendering

```{r}
library(here)
library(tidyverse)
library(spdep)

@@

car_black <- spautolm(`Death Cases` ~ `Percent Living in Poverty` + 
                        `Percent High School Education` + offset(Population),
                      listw = nablist_black_df, 
                      data = black,
                      family = "CAR")
print(summary(car_black))

@@

car_multiple <- spautolm(`Death Cases` ~ `Percent Living in Poverty` + 
                        `Percent High School Education` + offset(Population),
                         data = multiple,
                         listw = nablist_multiple_df,
                         family = "CAR")
print(summary(car_multiple))

@@

car_hispanic <- spautolm(`Death Cases` ~ `Percent Living in Poverty` + 
                        `Percent High School Education` + offset(Population),
                         data = hispanic,
                         listw = nablist_hispanic_df,
                         family = "CAR")
print(summary(car_hispanic))

@@

car_white <- spautolm(`Death Cases` ~ `Percent Living in Poverty` + 
                        `Percent High School Education` + offset(Population),
                      data = white,
                      listw = nablist_white_df,
                      family = "CAR")
print(summary(car_white))

>>>>

```{r}
library(here)
library(tidyverse)
library(spdep)
library(spatialreg)

@@

car_black <- spautolm(BOX ~ `Percent Living in Poverty` + 
                        `Percent High School Education`,
                      listw = nablist_black_df, 
                      data = black,
                      family = "CAR")
print(summary(car_black))

@@

car_multiple <- spautolm(BOX ~ `Percent Living in Poverty` + 
                        `Percent High School Education`,
                         data = multiple,
                         listw = nablist_multiple_df,
                         family = "CAR")
print(summary(car_multiple))

@@

car_hispanic <- spautolm(BOX ~ `Percent Living in Poverty` + 
                        `Percent High School Education`,
                         data = hispanic,
                         listw = nablist_hispanic_df,
                         family = "CAR")
print(summary(car_hispanic))

@@

car_white <- spautolm(BOX ~ `Percent Living in Poverty` + 
                        `Percent High School Education`,
                      data = white,
                      listw = nablist_white_df,
                      family = "CAR")
print(summary(car_white))

### 2023-04-08 code/analysis_code/Race/p5_mapping.Rmd
fixed rendering

Now I'm ready to start mapping! I'll begin by looking at the map of the raw HIV death rate across to contiguous US by selected race/ ethnicities.
```{r}
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Death Rate (estimated)`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "HIV Mortality Rate (raw) by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Deaths per 100,000") +
  scale_fill_distiller(palette = "RdYlGn")
```
The map for HIV-related death differs drastically across the United States by race/ethnicity. In most states, the highest HIV-related death were among Black person. This is especially true in the South region. Non-coastal western states consistently had lowest rates HIV-related deaths across the selected racial/ethnic groups, expect among multi-racial non-Hispanic persons. Non-Hispanic White persons were the only of the selected racial/ ethnic group that did not have a HIV-related death rate higher than ~10 per 100,000.

Next, let's map a probability map based on where we'd expect deaths if risk of HIV-related death was consistent across the US.
```{r}
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = Exp_DR)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race) +
  scale_fill_viridis_c() + theme_classic() + 
  theme(strip.text = element_text(size = 14),
    strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "NULL - Expected HIV Mortality Rate by Race/ Ethnicity
       in the United States, 2020", x = "", y = "", fill = "Rate")
```
Well.. I probably need to go back and reconsider using making a lambda based on specific race/ ethnic. I think that at this point, however, it is still fair to consider mapping an observed/ expected ratio using the expected deaths calculate from the lambda for the entire population. This will allow us to see differences across the race/ ethnicity, rather than masking disparities in rates due higher expected cases among certain high-risk populations, which would thus make the ratio closer to the null.

Let's map the observed/ expected ratio.
```{r}
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = OE_mortality_ratio)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race) +
  scale_fill_viridis_c() + theme_classic() + 
  theme(strip.text = element_text(size = 14),
    strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "Observed/ Expected HIV Mortality Ratio by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Ratio")
```
We can see that non-Hispanic Black persons had ratios consistently (much) greater than 1, suggesting a high risk among this population.

Next, let's using Dykes and Unwin's method of smooth map using a chi square value. 
```{r}
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = dykes_chi)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + theme_classic() + 
  theme(strip.text = element_text(size = 14),
        strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "Dyke's Chi Square by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Dykes and Unwin's Chi Square") +
  scale_fill_distiller(palette = "RdYlGn")
```
##Dykes and Unwin’s chi-square statistic
Dykes and Unwin (Dykes and Unwin 2001) technique  has the effect of creating a distribution of values closer to normal (as opposed to a Poisson distribution of rates and counts encountered thus far). We can therefore apply a diverging color scheme where green hues represent less than expected rates and red hues represent greater than expected rates.

https://mgimond.github.io/Spatial/mapping-rates-in-r.html
```{r}
E.Best <- EBest(df_sf$`Death Cases`,df_sf$Population)
df_sf$EBest <- E.Best$estmm * 100000
library(classInt)
brks1 <- classIntervals(df_sf$EBest, n = 10, style = "quantile")

library(RColorBrewer)
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Death Rate (estimated)`)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn")), 
                       breaks = round(brks1$brks)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom",
        legend.key.width = unit(1.5, "cm"), # Adjust the width of the legend
        legend.key.height = unit(0.5, "cm")) + # Adjust the height of the legend
  guides(fill = guide_colorbar(barwidth = 50, barheight = 1)) # Adjust the size of the colorbar
```

Next, let's make a map using the fitted values from the CAR model, the best performing model.
```{r}
##Smoothed map using fitted values.
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = fitted_values)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") 
```

Next, let's look at the map for the ratio of predicted to observed (P/O) values. 
```{r}
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = predicted_observed_risk)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") 
```

Interesting... While the AIC suggest that the CAR model performs well, the model has negative predicted values. This is problematic and suggests that it is not well-suited to the data.
Regardless, a higher P/O ratio (more red values) indicates that the model overestimated the observed values, while a lower P/O ratio (more green values) indicates that the model underestimates the observed values.

>>>>

Now I'm ready to start mapping! I'll begin by looking at the map of the raw HIV death rate across to contiguous US by selected race/ ethnicity.
```{r}
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Death Rate (estimated)`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "HIV Mortality Rate (raw) by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Death rate per 100,000") +
  scale_fill_distiller(palette = "RdYlGn") +
  theme(plot.title= element_text(hjust = 0.5))

ggsave(here("results/raw_rate_map.png"), units="in", width=10, height=7, dpi=300)
```
The map for HIV-related death differs drastically across the United States by race/ethnicity. In most states, the highest HIV-related death were among Black person. This is especially true in the South region. Non-coastal western states consistently had lowest rates HIV-related deaths across the selected racial/ethnic groups, expect among multi-racial non-Hispanic persons. Non-Hispanic White persons were the only of the selected racial/ ethnic group that did not have a HIV-related death rate higher than ~10 per 100,000.

Let's map the observed/ expected ratio.
```{r}
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = OE_mortality_ratio)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race) +
  scale_fill_distiller(palette = "RdYlGn") + theme_void() + 
  theme(strip.text = element_text(size = 14),
    strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "Observed/ Expected HIV Mortality Ratio by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))

ggsave(here("results/roe_map.png"), units="in", width=10, height=7, dpi=300)
```
We can see that non-Hispanic Black persons had ratios consistently (much) greater than 1, suggesting a high risk among this population.

Next, let's map after the the Box_Cox transformation of the ratio of observed/ expected cases
```{r}
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = BOX)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race) +
  scale_fill_distiller(palette = "RdYlGn") + theme_void() + 
  theme(strip.text = element_text(size = 14),
    strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "Ratio of Observed/ Expected Cases (Box Cox transformed) by Race/ Ethnicity
in the United States, 2020", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))
ggsave(here("results/roe_box_map.png"), units="in", width=10, height=7, dpi=300)
```

Next, let's map a probability map based on where we'd expect deaths if risk of HIV-related death was consistent across the US.
```{r}
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = Exp_DR)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race) +
  scale_fill_viridis_c() + theme_void() + 
  theme(strip.text = element_text(size = 14),
    strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "NULL - Expected HIV Mortality Rate by Race/ Ethnicity
       in the United States, 2020", x = "", y = "", fill = "Rate") +
  theme(plot.title= element_text(hjust = 0.5))
```
Well.. I probably need to go back and reconsider using making a lambda based on specific race/ ethnic. I think that at this point, however, it is still fair to consider mapping an observed/ expected ratio using the expected deaths calculate from the lambda for the entire population. This will allow us to see differences across the race/ ethnicity, rather than masking disparities in rates due higher expected cases among certain high-risk populations, which would thus make the ratio closer to the null.

Next, let's using Dykes and Unwin's method of smooth map using a chi square value. 
```{r}
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = dykes_chi)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + theme_void() + 
  theme(strip.text = element_text(size = 14),
        strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "Dykes and Unwin's Chi Square by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Chi Square") +
  scale_fill_distiller(palette = "RdYlGn") +
  theme(plot.title= element_text(hjust = 0.5))

ggsave(here("results/chi_map.png"), units="in", width=10, height=7, dpi=300)
```
##Dykes and Unwin’s chi-square statistic
Dykes and Unwin (Dykes and Unwin 2001) technique  has the effect of creating a distribution of values closer to normal (as opposed to a Poisson distribution of rates and counts encountered thus far). We can therefore apply a diverging color scheme where green hues represent less than expected rates and red hues represent greater than expected rates.

Next, let's make a map using the fitted values from the CAR model, the best performing model.
```{r}
##Smoothed map using fitted values.
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = fitted_values)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Smoothed Map of Square by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))

ggsave(here("results/fitted_map.png"), units="in", width=10, height=7, dpi=300)

```

Next, let's look at the map for the ratio of predicted to observed (P/O) values. 
```{r}
ggplot(df_sf, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = predicted_observed_risk)) +
  scale_fill_gradientn(colors = rev(brewer.pal(9, "RdYlGn"))) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Map of Ratio of Predicted to Observed Cases by Race/ Ethnicity in the United States, 2020", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))
``` 
We can see that the model fitted pretty well. A higher P/O ratio (more red values) indicates that the model overestimated the observed values, while a lower P/O ratio (more green values) indicates that the model underestimates the observed values. We can see that the model underestimated rates primarily in the non-coast north-western states and north central states for persons of Hispanic, non-Hispanic White, and non-Hispanic multiple racial/ ethnic backgrounds.

### 2023-04-08 products/manuscript/MANU.qmd
fixed rendering

---
title: "Manuscript Template for a Data Analysis Project"
subtitle: "Spatial Investigation and Exploration of Factors influencing HIV-related Mortality Rates in the United States"
author: Hayley Hemme
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../dataanalysis_template_references.bib
csl: ../apa.csl
---

@@

## Data aquisition

@@

*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

@@

@@

>>>>

---
title: "Spatial Investigation and Exploration of Factors influencing HIV-related Mortality"
subtitle: "United States, 2020"
author: Hayley Hemme
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../dataanalysis_references.bib
csl: ../apa.csl
---

@@

## Data acquisition

@@

![Histogram of the distribution of the ratio for observed/ expected cases(raw)](images/raw_roe_hist.png){width="300"}

![Histogram of the distribution of ratio of observed/ expected mortality after Box-Cox transformation](images/box_hist-01.png){width="300"}

![Diagnostics for Box Cox Transformed Ratio Observed/ Expected](images/box_diag.png){width="300"}*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

@@

|                                                     |                                                     |
|-----------------------|------------------------|
| ![Black](images/moran_black-01.png){width="200"}    | ![Multiple](images/moran_multiple.png){width="200"} |
| ![Hispanic](images/moran_hispanic.png){width="200"} | ![White](images/moran_white.png){width="200"}       |

@@

|                                              |                                         |
|-----------------------------|------------------------------|
| ![](images/raw_rate_map-02.png){width="250"} | ![](images/roe_map-03.png){width="250"} |
| ![](images/roe_box_map-01.png){width="250"}  | ![](images/chi_map-02.png){width="250"} |
| ![](images/fitted_map-02.png){width="250"}   |                                         |
